---
title: "Robin's Data Processing Example"
author: "Jeanette Clark"
date: "3/15/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning  = FALSE, message = FALSE}
library(tidyverse)
library(stringr)
```

Create a test dataframe, have to do some brute force on the column names to make sure they look like the real data.
```{r}
test <- data.frame(id = 1, "400.123" = 1, "512" = 1, "400.123_BR" = 1, "512_BR" = 1)
colnames(test)[2] <- "400.123"; colnames(test)[3] <- "512"
colnames(test)[4] <- "400.123_BR"; colnames(test)[5] <- "512_BR"
head(test)
```

Here we use some complex logic to determine which columns look like numbers. We try to coerce the column names to a number, and look for the column indices for ones that did not get turned into `NA`.
```{r}
i <- which(
      is.na(
       as.numeric(
        colnames(test))) == F)
```
Now we actually rename the column names of these numeric columns (the index of which is stored in `i`). We do this using `paste`, and we append a lower case x before these columns.
```{r}
colnames(test)[i] <- paste("x",colnames(test)[i], sep = "")
colnames(test)
```

Next, select the "raw" wavelength columns by selecting all columns that don't end in BR. This ensures all of the identifying columns are also selected.
```{r}
test_raw_wide <- test %>% select(-ends_with("_BR"))
print(test_raw_wide)
```
If we want to make this long, use gather over the wavelength (`starts_with("x")`) columns, and then use mutate to get rid of the "x" in the new wavelength column.
```{r}
test_raw_long <- test_raw_wide %>% 
  gather(key = "wavelength", value = "number_raw", starts_with("x")) %>%
  mutate(wavelength = str_replace_all(wavelength, "x", ""))
print(test_raw_long)

```
Now we do the same for the brightness adjusted columns, selecting all the columns except the ones that start with x.
```{r}
test_BR_wide <- test %>% select(-starts_with("x"))
print(test_BR_wide)
```
And finally, use gather and mutate to put this into long form and clean up the wavelength column.
```{r}
test_BR_long <- test_BR_wide %>% 
  gather(key = "wavelength", value = "number_BR", ends_with("_BR")) %>%
  mutate(wavelength = str_replace_all(wavelength, "_BR", ""))
print(test_BR_long)
```
